Automated Verification for Real-Time Systems using Implicit Clocks and an Extended Antimirov Algorithm

### 2022-04-16 paper/tv.tex
m

$[FV\text{-}\m{Value}]$  obtains the next program state by 
inheriting the current program state.
$[FV\text{-}Guard]$ computes the effects of \code{e} after concatenating
\code{(b{=}True)}? in to the current program states.  
{{\small\begin{gather*}
  \code{\frac{
  }
{  \vdash \{  \Pi, \Theta \}\  v \ 
\{ \Pi, \Theta \}} \  
 {[FV\text{-}Value]}} 
 \qquad
 \code{\frac{
  \vdash \{  \Pi, \Theta \cdot (b{=}True)? \}\  e \ 
  \{ \Pi^\prime, \Theta^\prime \}
  }
{  \vdash \{  \Pi, \Theta \}\  [b]\ e \ 
\{  \Pi^\prime, \Theta^\prime \}} \  
 {[FV\text{-}Guard]}} 
\end{gather*}}}

@@

  \end{tabular}
  \end{table}

@@

To show the feasibility of our approach, we have prototyped 
our automated verification system 
using OCaml. The proof obligations generated by the verifier 
are discharged using constraint solver Z3 
\citep{DBLP:conf/tacas/MouraB08}. 
We prove termination and soundness of the TRS. We validate 
the front-end forward verifier against the state-of-the-art PAT 
\cite{DBLP:conf/cav/SunLDP09} model checker for conformance. 
Next, we show experimental results and case studies to evaluate the 
capability of our proposals.

@@

We conduct experiments on a MacBook Pro with a 2.6 GHz Intel Core i7 processor. Given our benchmark, the running time varying from 0 to 500 ms.

@@

\autoref{tab:Experimental} presents the evaluation results of a microbenchmark. % summary of a lightweight . 
In this experiment, we select 16 \timedL\ programs, and  annotate temporal specifications
with a 1:1 ratio for succeeded/failed cases. 
It records: 
the index of the program, \textbf{No.};
lines of code, \textbf{LOC}; 
effects inference time, \textbf{Infer(ms)}; 
number of valid properties, \textbf{$\mysharp$Prop(\cmark)}; 
average proving time for the valid properties, \textbf{Avg\text{-}Prove(ms)};
number of invalid properties, \textbf{$\mysharp$Prop(\xmark)}; and 
average disproving time for the invalid properties, \textbf{Avg\text{-}Dis(ms)}. 
Times are counted using milliseconds, and the experiment is done on 
a MacBook Pro with a 2.6 GHz 6-Core Intel Core i7 processor.

\begin{table}[!ht]
\caption{\label{tab:Experimental} Experimental Results. }
%\vspace{-1mm}
\centering
\renewcommand{\arraystretch}{1.1}
\begin{tabular} {c|c|c|c|c |c|c|c}
\hline
\textbf{No.} & \textbf{LOC} & \textbf{Infer(ms)} & \textbf{$\mysharp$Prop(\cmark)}   
& \textbf{Avg\text{-}Prove(ms)} & \textbf{$\mysharp$Prop(\xmark)}   & \textbf{Avg\text{-}Dis(ms)} &\textbf{$\mysharp$AskZ3} \\ \hline\hline
1    &    30    & 0.031  &  5     &  4.185   & 5   &  3.242  & 68 \\ \hline
2    &    35    & 0.038  &  3     &  13.761  & 7   &  9.331  & 214 \\ \hline
3    &    44    & 0.049  &  4     &  12.149  & 6   & 7.853   & 215 \\ \hline
4    &    72    & 0.061  &  5     &  37.268 & 5   & 28.3054 & 617 \\ \hline
5    &    95    & 0.105  &  7     &  6.139    & 7   &  5.238  & 186 \\ \hline
6    &    108   & 0.16   &  7     &  43.667    & 7   &  17.178  & 818 \\ \hline
7    &    106   & 0.142  &  7     &  18.776    & 7   &  8.432  & 346 \\ \hline
8    &    144   & 0.141  &  7     &  22.238    & 7   &  10.051  & 426 \\ \hline
9    &    155   & 0.308  &  9     &  194.666   & 9   &  56.344  & 4059 \\ \hline
10   &    174   & 0.373  &  9     &  78.132   & 9   &   24.665  & 1740 \\ \hline
11   &    196   & 0.557  &  9     &  111.100   & 9   &   31.262  & 2403 \\ \hline
12   &    213   & 0.516  &  9     &  114.667   & 9   &  30.284   & 2484 \\ \hline
13   &    265   & 1.895  &  11    & 616.619    & 11  &  302.975  & 9068 \\ \hline
14   &    271   & 1.702  &  11    &  302.145   & 11  &  131.938  & 7779 \\ \hline     
15   &    276   & 2.121 & 11     &  466.174  & 11  &   181.116  & 12239 \\ \hline
16   &    314   & 1.654 &  11    & 1094.231   & 11  &  465.704  & 32620 \\ \hline            
%\hline
%\textbf{\quad \ \ Total}  & 2546    &  235   &  446.88   &   305.33   \\ \hline
\end{tabular}
%  \vspace{-1mm}
\end{table}

@@

As shown in \tabref{tab:reoccur}, the LHS formula is a 
disjunction of two effects. An inclusion with a disjunctive 
antecedent succeeds if both disjunctions entail the consequent, 
shown in step {\textcircled{1}}.  
%Now we have 2 subtrees, the first subtree is for
%  \code{
%    \textcolor {darkred}{t_L{<}3} : \{A\}^\star \mysharp  t_L \cdot \{B\}
%    \CONTAIN
%    \textcolor {darkred}{t_R{<}4} : \{A\}^\star \mysharp  t_R \cdot \{B\}
%  }
%and the second one is for
%  \code{
%    \textcolor {darkred}{True} : {B} \CONTAIN
%    \textcolor {darkred}{t_R{<}4} : \{A\}^\star \mysharp  t_R \cdot \{B\}
%  }.
%In order to prove the original proposition, both of the two new propositions have
%to be proved.
%Using the disjunction rule \code{\effect_1 \vee \effect_2 \CONTAIN \effect \iff
%\effect_1 \CONTAIN \effect \ and \  \effect_2 \CONTAIN \effect}, we split the proposition
%in . 
In step \textcircled{2} of the first sub-tree, in order to eliminate the first event \code{\{B\}}, \code{\{A\}^\star \mysharp  t_R} has to be reduced to be \code{\epsilon}, therefore the time constraints have nothing to do in this branch. 
% is 
%eliminated by \code{\epsilon}, 
Then in steps \textcircled{4}, we
can simply prove it with the side constraint entailment \code{\textcolor {darkred}{True} \Rightarrow  \textcolor {darkred}{True}} succeed.
%and \code{\epsilon \CONTAIN \epsilon}.

Looking at the second sub-tree, in step \textcircled{5}, \code{t_L} and \code{t_R} are firstly split 
into \code{t_L^1{+}t_L^2} and \code{t_R^1{+}t_R^2}. 
Then in step {\textcircled{6}, 
\code{\{A\} \mysharp  t_L^1} together with \code{\{A\} \mysharp  t_R^1} are eliminated, unifying \code{t_L^1} and \code{t_R^1} (adding the constraint \code{ t_L^1 {=}  t_R^1}).
%, also
%are unified as new constraint \code{t_L^1{=}t_R^1}.
In step {\textcircled{8}, we observe the proposition is isomorphic with one of the the previous step, marked using \code{\textcolor{darklavender}{(\ddagger)}}. 
%one in
%step \code{\textcircled{1}}, 
Hence we apply the rule \code{[REOCCUR]} to prove it with a succeed side constraints entailment: \code{\textcolor {
      darkred}{t_L^1{+}t_L^2{=}t_L \wedge t_L{<}3 \wedge t_R{=}t_R^1{+}t_R^2 \wedge t_L^1{=}t_R^1
      \wedge t_L^2{=}t_R^2
    }\Rightarrow
    \textcolor {darkred}{t_R{<}4} }. 
%Since the structure of terms \code{t_L^2, t_L} and \code{t_R^2, t_R} are isomorphic,
%we will need to unify \code{t_L^2} and \code{t_R^2} in order to get enough
%time constraints. And in step \code{\textcircled{5}} we prove the proposition to
%be true.

@@

@@

Classical LTL
extended propositional logic with the temporal operators 
\code{\mathcal{G} } (``globally") 
and 
 \code{\mathcal{F}} (``in the future"), which we also write
 \code{\square} and \code{\Diamond}, respectively. 
 %and introduced the concept of fairness. %, which ensures an infinite-paths semantics. 
 LTL was subsequently extended to include the  \code{\mathcal{U} }
  (``until") and  \code{\mathcal{X}} (``next time") 
  operators. 
  Our \timedEffects\ draws similarities to the metric temporal logic (MTL), 
which is defined similarly to LTL, 
  where a set of non-negative real number is added to temporal modal operators.   
  As shown in 
  \autoref{tab:ltl_effects}, we are able to recursively encode 
  these basic MTL operators into \timedEffects, making it 
   more intuitive and readable, mainly when nested operators 
   occur.

@@

\subsubsection{Beyond Regular, Context-Free and Context-Sensitive}

The paradigmatic non-regular linear language: 
\code{t{>} 0 \wedge \anyevent{A}\mysharp{t}\cdot\anyevent{B}\mysharp{t}}, 
can be naturally expressed by the depended effects. 
Besides, the effects can also express grammars such as \code{t{>} 0 
\wedge \anyevent{A}\mysharp{t}\cdot\anyevent{B}\mysharp{t}
\cdot\anyevent{C}\mysharp{t}}, or 
\code{n{>} 0 \wedge m{>} 0 \wedge  \anyevent{A}\mysharp{n}
\cdot\anyevent{B}\mysharp{m}\cdot\anyevent{C}\mysharp{n}}, 
which are beyond context-free grammar.
%(by a contradiction proof using Pumping lemma) 
%and categorized as context-sensitive grammar. 
Those examples show that the traces which cannot be recognized 
even by push-down automata (PDA) can be represented by our effects. 
%As regular grammar and context-free grammar are equivalent to finite state automata (FSA) and  respectively, our effects can not be fully translated into either FSA or PDA.
However, such specifications are significant, suppose we 
have a traffic light control system, we could have a 
specifications \code{n{>} 0 \wedge m{>} 0 \wedge  
(\anyevent{Red}\mysharp{n}\cdot\anyevent{Yellow}\mysharp{m}\cdot
\anyevent{Green}\mysharp{n})^\star}, which specifies that  
(i) all the colors will occur at each life circle; and 
(ii) the duration of the green light and the red light is 
always the same. Moreover, these effects can not be translated 
into linear bounded automata (LBA) either, which equivalents to 
context-sensitive grammar, as LBA are only capable of expressing 
finite traces.

@@

As the examples show, our proposed effects logic and the abstract semantics for \code{\timedL} not only tightly capture the behaviors of a mixed synchronous and asynchronous concurrency model but also help to mitigate the programming challenges in each paradigm. Meanwhile, the inferred temporal traces from a given reactive program enable a compositional timed temporal verification at the source level, which is not yet supported by existing temporal verification techniques.

Works based on such a TRS \cite{DBLP:conf/icfem/SongC20,DBLP:conf/vmcai/SongC21,DBLP:journals/tcs/AntimirovM95,DBLP:journals/ijfcs/AlmeidaMR09,DBLP:conf/fsttcs/KeilT14,DBLP:journals/jcss/Hovland12}  show its feasibility and suggest that this method is a better average-case algorithm than those based on the comparison of automata.

>>>>

\vspace{1mm}
$[FV\text{-}\m{Value}]$  obtains the next program state by 
inheriting the current program state.
$[FV\text{-}Guard]$ computes the effects of \code{e} after concatenating
\code{(b{=}True)}? in to the current program states.  
{{\small\begin{gather*}
  \code{\frac{
  }
{  \vdash \{  \Pi, \Theta \}\  v \ 
\{ \Pi, \Theta \}} \  
 {[FV\text{-}Value]}} 
 \qquad
 \code{\frac{
  \vdash \{  \Pi, \Theta \cdot (b{=}True)? \}\  e \ 
  \{ \Pi^\prime, \Theta^\prime \}
  }
{  \vdash \{  \Pi, \Theta \}\  [b]\ e \ 
\{  \Pi^\prime, \Theta^\prime \}} \  
 {[FV\text{-}Guard]}} 
\end{gather*}}}

@@

  \end{tabular}
  \vspace{-2mm}

  \end{table}

@@

To show the feasibility of our approach, we have prototyped 
our automated verification system 
using OCaml. The proof obligations generated by the verifier 
are discharged using constraint solver Z3 
\citep{DBLP:conf/tacas/MouraB08}. 
We prove termination and soundness of the TRS. We validate 
the front-end forward verifier against the state-of-the-art PAT 
\cite{DBLP:conf/cav/SunLDP09} model checker for conformance. 
Next, we show experimental results and case studies to evaluate our proposals.

@@

@@

\autoref{tab:Experimental} presents the evaluation results of a microbenchmark. % summary of a lightweight .

\begin{table}[!ht]
\caption{\label{tab:Experimental} Experimental Results. }
\vspace{-2mm}
\centering
\renewcommand{\arraystretch}{1}
\begin{tabular} {c|c|c|c|c |c|c|c}
\hline
\textbf{No.} & \textbf{LOC} & \textbf{Infer(ms)} & \textbf{$\mysharp$Prop(\cmark)}   
& \textbf{Avg\text{-}Prove(ms)} & \textbf{$\mysharp$Prop(\xmark)}   & \textbf{Avg\text{-}Dis(ms)} &\textbf{$\mysharp$AskZ3} \\ \hline\hline
1    &    30    & 0.031  &  5     &  4.185   & 5   &  3.242  & 68 \\ \hline
2    &    35    & 0.038  &  3     &  13.761  & 7   &  9.331  & 214 \\ \hline
3    &    44    & 0.049  &  4     &  12.149  & 6   & 7.853   & 215 \\ \hline
4    &    72    & 0.061  &  5     &  37.268 & 5   & 28.3054 & 617 \\ \hline
5    &    95    & 0.105  &  7     &  6.139    & 7   &  5.238  & 186 \\ \hline
6    &    108   & 0.16   &  7     &  43.667    & 7   &  17.178  & 818 \\ \hline
7    &    106   & 0.142  &  7     &  18.776    & 7   &  8.432  & 346 \\ \hline
8    &    144   & 0.141  &  7     &  22.238    & 7   &  10.051  & 426 \\ \hline
9    &    155   & 0.308  &  9     &  194.666   & 9   &  56.344  & 4059 \\ \hline
10   &    174   & 0.373  &  9     &  78.132   & 9   &   24.665  & 1740 \\ \hline
11   &    196   & 0.557  &  9     &  111.100   & 9   &   31.262  & 2403 \\ \hline
12   &    213   & 0.516  &  9     &  114.667   & 9   &  30.284   & 2484 \\ \hline
13   &    265   & 1.895  &  11    & 616.619    & 11  &  302.975  & 9068 \\ \hline
14   &    271   & 1.702  &  11    &  302.145   & 11  &  131.938  & 7779 \\ \hline     
15   &    276   & 2.121 & 11     &  466.174  & 11  &   181.116  & 12239 \\ \hline
16   &    314   & 1.654 &  11    & 1094.231   & 11  &  465.704  & 32620 \\ \hline            
%\hline
%\textbf{\quad \ \ Total}  & 2546    &  235   &  446.88   &   305.33   \\ \hline
\end{tabular}
%  \vspace{-1mm}
\end{table}

In this experiment, we select 16 \timedL\ programs, and  annotate temporal specifications
with a roughly 1:1 ratio for succeeded/failed cases. 
Times are counted using milliseconds, and the experiment is done on 
a MacBook Pro with a 2.6 GHz 6-Core Intel Core i7 processor. 
The table records: 
\textbf{No.}, the index of the program;
\textbf{LOC}, lines of code; 
\textbf{Infer(ms)}, effects inference time; 
\textbf{$\mysharp$Prop(\cmark)}, number of valid properties; 
\textbf{Avg\text{-}Prove(ms)}, average proving time for the valid properties;
\textbf{$\mysharp$Prop(\xmark)}, number of invalid properties; 
\textbf{Avg\text{-}Dis(ms)}, average disproving time for the invalid properties; and 
\textbf{$\mysharp$AskZ3}, number of querying Z3 solver through out the experiments.

\paragraph{Discussion:} Generally, the inference time and proving 
time increase while the program's length increases. 
Besides, while \emph{the number of querying Z3 per property} (\textbf{$\mysharp$AskZ3}/(\textbf{$\mysharp$Prop(\cmark)} +\textbf{$\mysharp$Prop(\xmark)})) 
goes up, the 
proving/disproving time goes up.
Furthermore, we notice that 
%that no matter how long the programs are, 
the disproving
times for invalid properties are constantly lower than the proving process, 
regardless of the program complexity. This finding echos the 
insights from prior TRS-based works \cite{DBLP:conf/icfem/SongC20,DBLP:journals/tcs/AntimirovM95,DBLP:journals/ijfcs/AlmeidaMR09,DBLP:conf/fsttcs/KeilT14,DBLP:journals/jcss/Hovland12}, 
which suggest that TRS is a better average-case algorithm than 
those based on the comparison of automata.

We summary: a TRS is efficient because 
\emph{it only constructs automata as far as it needs}, which makes 
it more efficient when disproving incorrect specifications, as we can 
disprove it earlier without constructing the whole automata. In other 
words, the more incorrect specifications are, the more efficient our solver is.

@@

@@

As shown in \tabref{tab:reoccur}, the LHS formula is a 
disjunction of two effects. An inclusion with a disjunctive 
antecedent succeeds if both disjunctions entail the consequent, 
shown in step {\textcircled{1}}.

In step \textcircled{2} of the first sub-tree, in order to eliminate the first event \code{  B  }, \code{  A  ^\star \mysharp  t_R} has to be reduced to be \code{\epsilon}, therefore the time constraints have nothing to do in this branch. 
Then in steps \textcircled{4}, we
can simply prove it with the side constraint entailment \code{\textcolor {darkred}{True} \Rightarrow  \textcolor {darkred}{True}} succeed.

Looking at the second sub-tree, in step \textcircled{5}, \code{t_L} and \code{t_R} are firstly split 
into \code{t_L^1{+}t_L^2} and \code{t_R^1{+}t_R^2}. 
Then in step {\textcircled{6}, 
\code{  A   \mysharp  t_L^1} together with \code{  A   \mysharp  t_R^1} are eliminated, unifying \code{t_L^1} and \code{t_R^1} (adding the constraint \code{ t_L^1 {=}  t_R^1}).
In step {\textcircled{8}, we observe the proposition is isomorphic with one of the the previous step, marked using \code{\textcolor{darklavender}{(\ddagger)}}. 
%one in
%step \code{\textcircled{1}}, 
Hence we apply the rule \code{[REOCCUR]} to prove it with a succeed side constraints entailment: \code{\textcolor {
      darkred}{t_L^1{+}t_L^2{=}t_L \wedge t_L{<}3 \wedge t_R{=}t_R^1{+}t_R^2 \wedge t_L^1{=}t_R^1
      \wedge t_L^2{=}t_R^2
    }\Rightarrow
    \textcolor {darkred}{t_R{<}4} }.

@@

\begin{comment}
  Classical LTL
extended propositional logic with the temporal operators 
\code{\mathcal{G} } (``globally") 
and 
 \code{\mathcal{F}} (``in the future"), which we also write
 \code{\square} and \code{\Diamond}, respectively. 
 %and introduced the concept of fairness. %, which ensures an infinite-paths semantics. 
 LTL was subsequently extended to include the  \code{\mathcal{U} }
  (``until") and  \code{\mathcal{X}} (``next time") 
  operators. 
  Our \timedEffects\ draws similarities to the metric temporal logic (MTL), 
which is defined similarly to LTL, 
  where a set of non-negative real number is added to temporal modal operators.   
  As shown in 
  \autoref{tab:ltl_effects}, we are able to recursively encode 
  these basic MTL operators into \timedEffects, making it 
   more intuitive and readable, mainly when nested operators 
   occur.

@@

\subsubsection{Beyond Regular, Context-Free and Context-Sensitive}
\end{comment}

The paradigmatic non-regular linear language: 
\code{t{>} 0 \wedge \anyevent{A}\mysharp{t}\cdot\anyevent{B}\mysharp{t}}, 
can be naturally expressed by the depended effects. 
Besides, the effects can also express grammars such as \code{t{>} 0 
\wedge \anyevent{A}\mysharp{t}\cdot\anyevent{B}\mysharp{t}
\cdot\anyevent{C}\mysharp{t}}, or 
\code{n{>} 0 \wedge m{>} 0 \wedge  \anyevent{A}\mysharp{n}
\cdot\anyevent{B}\mysharp{m}\cdot\anyevent{C}\mysharp{n}}, 
which are beyond context-free grammar.
%(by a contradiction proof using Pumping lemma) 
%and categorized as context-sensitive grammar. 
Those examples show that the traces which cannot be recognized 
even by push-down automata (PDA) can be represented by our effects. 
%As regular grammar and context-free grammar are equivalent to finite state automata (FSA) and  respectively, our effects can not be fully translated into either FSA or PDA.
However, such specifications are significant, suppose we 
have a traffic light control system, we could have a 
specifications \code{n{>} 0 \wedge m{>} 0 \wedge  
(\anyevent{Red}\mysharp{n}\cdot\anyevent{Yellow}\mysharp{m}}
\code{\cdot \anyevent{Green}\mysharp{n})^\star}, which specifies that  
(i) all the colors will occur at each life circle; and 
(ii) the duration of the green light and the red light is 
always the same. Moreover, these effects can not be translated 
into linear bounded automata (LBA) either, which equivalents to 
context-sensitive grammar, as LBA are only capable of expressing 
finite traces.

@@

As the examples show, our proposed effects logic and the abstract semantics for 
\code{\timedL} not only tightly capture the behaviors of a time-critical model 
but also help to increase the capability of existing techniques. 
Meanwhile, the inferred temporal traces from a given reactive program enable a compositional 
timed temporal verification at the source level, which is not yet supported by existing temporal 
verification techniques.
